---
title: "Election Prediction Analysis: Trump vs. Harris"
subtitle: "Assessing the Probability of Electoral College Outcomes Based on Polling Data"
author: 
  - Harsh Pareek
  - Benji Fleurence
  - Arshh Relan
thanks: "Code and data are available at: https://github.com/HarshMPareek/USA_Election_Prediction
date: "`r Sys.Date()`"
abstract: "This paper aims to predict the outcome of the 2024 U.S. Presidential Election between Donald Trump and Kamala Harris by analyzing aggregated state-level polling data. Using a linear regression model and Monte Carlo simulations, we estimate the probability of each candidate winning the Electoral College."
output: pdf_document
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```


# Introduction

Over the last few election cycles, pollsters' reputations have taken a hit. In back-to-back general elections, pollsters predicted massive Trump losses, only for President Trump to win in 2016 and narrowly lose in 2020. Pollsters are keen on correcting their previous errors. For political campaigns that have already exceeded $16 billion between the two candidates and an election decided by slim margins, polling is vital to assess where they should use their resources.

The estimand of our paper consists of the eligible voters in the United States. Our goal is to predict what the electorate will decide, focusing on specific sub-sections in more contested regions. This focus is crucial because of the Electoral College system the U.S. uses instead of a pure popular vote. In this system, when a president wins a state, they receive all of its electoral votes, and each state has a unique number of votes. Consequently, because certain states are more competitive than others, their value in our sample should be weighted higher. For instance, a respondent from Pennsylvania will have a heavier impact on the election outcome than someone from Oklahoma.

This paper employs a combination of linear regression modeling and Monte Carlo simulations to analyze aggregated polling data. We aim to provide a comprehensive prediction of the election outcome by considering various factors influencing voter behavior and polling accuracy.

The remainder of this paper is structured as follows. @sec-data provides a detailed overview of the dataset, including variable descriptions and exploratory data analysis. @sec-model discusses the modeling approach and justifications. @sec-results presents the findings from the analysis, and @sec-discussion interprets these results in the broader political context. Finally, @sec-appendix contains supplementary materials and detailed methodological information.


# Data {#sec-data}

## Overview

We utilize aggregated polling data from multiple sources across all 50 U.S. states and the District of Columbia. The dataset focuses exclusively on polls that include both Donald Trump and Kamala Harris, ensuring consistency and comparability in the analysis. This approach allows us to capture the competitive dynamics between the two candidates across various states with differing electoral weights.

##Broader Context


Polling data serves as a critical tool for predicting election outcomes by gauging voter intentions and preferences. By aggregating this data at the state level, we account for regional variations in political landscapes and voter behavior. This granularity is essential for accurately simulating Electoral College outcomes, where state-by-state victories determine the overall election result.

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Veriables explained

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 

## Predictor variables

Days Until Election
The days_until_election variable measures the number of days remaining until the election date at the time the poll was conducted. This temporal variable helps capture trends and shifts in voter sentiment as the election approaches.

Sample Size
Sample_size indicates the number of respondents in each poll. Larger sample sizes generally provide more reliable estimates of voter preferences and reduce the margin of error.

Methodology
The methodology variable categorizes the polling method used, such as live phone, online panel, or mixed methods. Different methodologies can introduce varying levels of bias and representativeness in the polling data.

Population
Population denotes the demographic segment targeted by the poll, including factors like age, gender, race, education level, and income bracket. This variable ensures that the polling data accurately reflects the diversity of the electorate.

# Model

The goal of our modeling strategy is twofold. Firstly, to quantify the influence of various predictors on polling percentages for each candidate. Secondly, to utilize these insights to simulate Electoral College outcomes through Monte Carlo simulations.

Here, we briefly describe the linear regression model used to investigate the relationship between polling percentages and predictors such as candidate name, days until election, sample size, methodology, population demographics, and state-specific factors. Detailed model specifications and diagnostics are included in @sec-model-details.

## Model set-up

## Linear Regression Model Specification

Define \( y_i \) as the normalized polling percentage for candidate \( i \). The linear regression model is specified as follows:

\[
y_i = \beta_0 + \beta_1 \cdot \text{candidate\_name} + \beta_2 \cdot \text{days\_until\_election} + \beta_3 \cdot \text{sample\_size} + \beta_4 \cdot \text{methodology} + \beta_5 \cdot \text{population} + \beta_6 \cdot \text{state} + \epsilon_i
\]

Where:

- \( \beta_0 \): The intercept.
- \( \beta_1 \): Captures the effect of being Kamala Harris relative to Donald Trump.
- \( \beta_2 \): Measures the change in polling percentage per additional day until the election.
- \( \beta_3 \): Represents the effect of sample size on polling percentage.
- \( \beta_4 \), \( \beta_5 \), and \( \beta_6 \): Account for variations due to polling methodology, targeted population, and state-specific factors, respectively.
- \( \epsilon_i \): The error term.

We implement the model using R's `lm()` function, leveraging the `dplyr` and `tidyr` packages for data manipulation. Model diagnostics were conducted to assess assumptions, including residual analysis and variance inflation factors (VIF) to check for multicollinearity.



### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## Have the Polls finally caught up to Trump? 

How voters turn out for Kamala Harris is yet to be seen, but the data on Trump is becoming pretty evident. The best predictor of future elections is past elections, and with Donald Trump, we now have two election cycles to go off of. In both, Trump vastly over-shot the mainstream poll’s projections, projected to lose big both times; in 2016, he overcame pre-election polls due to his unforeseen success with non-college-educated white men and women, and in 2020, he made gains in the female and Hispanic voter shares narrowly losing when he was projected to lose in a landslide. Trump found success by connecting with groups of voters who were not very politically engaged, did not always vote, and certainly did not complete pre-election surveys. One key difference between polls today and 2016 is accounting for education level. Trump found his “silent” voters by motivating mostly white non-college graduates, who were disproportionally likely to fill out a survey in comparison with college-educated voters, skewing the polls towards Clinton. Other changes pollsters are continuing to make are changing the weights given to an individual survey respondent based on some general characteristic such as race, gender, or education level, and some more specific attributes like whether you’ve donated blood recently or whether you rent or buy your home in an attempt to try and better represent the true voter population. What’s different about this year compared to the previous two campaigns in 2016 and 2020 is that Trump is in a statistical tie rather than heavily behind. For Trump and Republican strategists who are doubtful pollsters have finally figured out how to accurately represent Trump’s electorate, they perceive a tie to forecast a guaranteed win.

## Does Kamala Harris have a chance? 

There are a lot of factors that suggest Kamala Harris shouldn’t have had a shot at winning this race to begin with. President Biden dropped out of the race only 107 days before election day. This unprecedented move’s two closest examples in history were Truman and Johnson, who both dropped out with over 200 days to go, and both saw their replacements lose in the general election [@poynter_candidate]. In addition, ABC News’ most recent polls found that 74% of likely voters believe this country is headed in a wrong direction [@abcnews_election]. This negative sentiment has grown for the current administration due to their handling of global wars, ongoing border issues, and inflation, drawing harsh criticism from both sides of the aisle. Harris and Trump are clearly aware of this political atmosphere, as Trump has attempted to tie Harris to everything Biden and his administration have done. At the same time, Harris has urged voters to turn the page on the Trump era of politics. Whoever wins that narrative battle will most likely be the next president of the United States. 

So, although it is an uphill battle, there are reasons for Harris and her campaign to believe they have a chance in this election. For the Harris campaign, they hope pollsters have finally figured out Trump’s silent voter and, in doing so, may have overlooked some of her support. The largest area in which Harris believes she can gain on her margins is with female voters. While polls show Trump ahead in issues like the Economy and the Border, Harris’ counter is her lead on Abortion and women’s rights[@nbcnews_poll]. Because of this advantage, the Harris campaign hopes that women who may have previously said on surveys or calls in front of their husbands that they were voting for Trump will actually vote for Harris on election day. Her campaign also hopes young voters who claim to have abandoned the democratic party over their handling of the war in Gaza will ultimately side with her when faced with the two choices of her or Trump. And if Latino voters who are angered over recent comments made about Puerto Rico at the Trump rally in Madison Square Garden will halt the gains Trump has been making in the Latino community since 2016. While it’s unlikely all demographics perfectly go Harris’ way, in a race as close as this one, any gain on the margins could be what seals it for her.

## What do we think will happen? 

Both of these candidates are relying on sampling errors in the polls. Trump believes his voter delegation is untraceable by the polls, and ironically, so is Harris. However, are the tight margins we see throughout the polls a signal that Trump is going to run away with this election or have the polls finally figured out how to predict Trump's support accurately? I lean towards Harris squeaking out a victory by the thinnest of margins. When I look towards the Harris campaign, I see more potential. Women moving across the aisle due to concerns over abortion, a strong social media following boosting young voter support, or the Latino community turning against Trump halting the momentum he had going since 2016. One recent poll suggesting these trends may come to fruition is a new Iowa poll[@desmoinesregister_poll] done by respect pollster J.Ann Selzer, who has predicted Iowa both correctly and accurately since 2008. In this poll of 808 Iowans, Selzer and Company have Harris leading by a 3-point margin, withen the margin of error, but most other outlets project Trump to win by nearly 10 points as he has in the past two election cycles. Selzer points to a shift in the female and age demographics towards Harris as the reason for her projected victory. These are the types of gains the Harris campaign will have to make with multiple communities across America if she wants to become the next president of the United States, but according to this poll, those gains are there to be made. 

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}
## Appendx 1

One poll that was frequently sampled throughout our analysis was ABC News' poll[@abcnews_polling] completed by Langer Research Associates in partnership with Ipsos. Their most recent poll was created to predict the outcome of the general election. They sampled eligible voters across the United States randomly through addressed-based sampling. They sent invitations to 4,335 participants, 2,496 sent a response, and 47 were discarded due to not filling out over half the questions or completing the survey in the top 1% time, leaving a sample of 2,449 likely voters. Participantsants were sent an online surusing Ipsos Knowledge Panel vey to complete. Ipsos then weighs respondents based on race, ethnicity, sex, religion, education, region, household income, household size, marital status, home ownership, metropolitan area, as well as passed voting preference and preferred language. These weights measure the probability of portion-to-size sample selections, increasing participants' probability of selection if their sample weights are higher. Specifically, respondents between the ages of 18-29, black and Latino, are all oversampled to make up for lack of representation and handle non-response bias. 

Langer's research settled on a 95% confidence interval to calculate their sampling error, but note that other sources of error can arise from questionnaire wording, order, and systematic noncoverage or selection bias. While an online survey is cost-effective, far-reaching, and convenient both in recording and receiving data, there are some negatives. The response rate is low, below 50%, and sample bias as those who respond to surveys are more likely to vote Democrat (the trap pollsters fell into in 2016 and 2020), and you can't observe the environment the respondents answer the survey in to ensure no outside bias. However, Langer's research disregarded the concerns for their low response rate by citing research papers, including this one written in 2000: "Consequences of Reducing Non-response in a National Telephone Survey" (Keeter, Miller, Kohut, Groves & Presser, POQ 64:125-48)[@abcnews_polling], which found similar results between response rates of 36%-61%. 

## Apendix 2

Objective: The goal of this survey methodology is to develop a reliable, predictive model for the U.S. presidential election by surveying a representative sample of eligible American voters. This approach considers demographic, geographic, and political factors influencing voting behavior, with a budget of $100,000 to ensure comprehensive reach, quality data collection, and robust analysis.
Methodology Overview: Utilizing probability sampling techniques, data validation processes, and advanced poll aggregation, this document outlines a strategy designed to yield accurate and generalizable election forecasts.
 
Strategy: We will use stratified random sampling. This method involves dividing the target population into subgroups based on demographics such as age, gender, race, education, income, and geographic region. Random sampling within each subgroup ensures a representative cross-section of voters. We will compliment this with quota sampling to ensure that each demographic group meets sufficient representation, even if probability sampling proves challenging.
We will stratify based on age, gender, race, ethnicity, education level, household income bracket, geographical region of the USA. This allows for greater forecasting accuracy.
The ideal sample size would be 10,000 individuals. The analysis will adjust the weightage for each demographic to ensure there is proportional representation in the forecasting.
 
Recruitment Methodology: Targeted ads on social media browsers and emails based on demographics (age, gender, location, political interests). Collaborate with non-profits and civic organizations to improve reach and respondent diversity. Participation in the survey will also be incentivized, participants will have the chance to enter lucky draw competitions that have gift cards as prizes.
 
The survey will be created and distributed using Qualtrics for enhanced security and functionality.
Questions in the survey will progress in the following broad sections. At first, demographic based questions, secondly questions on political views and intended candidate, lastly verification. Surveys should not take more than 5 minutes to complete.
 
Sample budget allocation:
Outreach: $80,000
Incentives: $10,000
Qualtrics and data analysis software: $10,000
 
A sample survey is attached here.

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


